---
title: "Bellabeat Fitness Data Analysis"
author: "Uju Iloabachie"
date: "`r format(Sys.time(), '%B %d %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1.INTRODUCTION

Bellabeat, is a high-tech company that  manufactures health-focused smart products for women such as Bellabeat app, leaf, Time and Spring.
These products connect to app to track activity, sleep, stress and hydration levels. Bellabeat also offers subscription-based membership program 
for users for personalized guidance on nutrition, activity, sleep, health, beauty and mindfulness based on their lifestyle goals.

Bellabeat is a successful small company and founded by Urska Srsen and Sando Mur. Srsen believes that analysis of Bellabeat's available consumer 
data would reveal more opportunities for growth and become a larger player in the global smart device market.


## 2. ASK

### 2.1 Identify the business task
The business task is to gain insight on how consumers use non-Bellabeat smart devices and to provide recommendation for how trends in the smart
device usage data can inform Bellabeat marketing strategy for opportunity for growth.

### 2.2 Consider Key Stakeholders
The key stakeholders include
* Urska Srsen (Co-founder and Chief Creative Officer)
* Sando Mur (Co-founder and Mathematician)
* Marketing analytics team


## 3. PREPARE

### 3.1 Data Source and Organization
The data used for this analysis is the FitBit Fitness Tracker data set obtained from [Kaggle.](https://www.kaggle.com/datasets/arashnic/fitbit?resource=download). 
The data was made available through [Mobius](https://www.kaggle.com/arashnic). The data is downloaded and  stored locally.

The FitBit Fitness tracker data zip has 18 csv files. 15 of which are presented in long format and 3 in wide format.It contains personal information of 
of eligible fitbit users. These information include physical activity, amount of sleep, weight, steps, heart rate, calories intake etc.

### 3.2 Data Credibility
The data contains information of women who were eligible and consented to give their health information as oppose to every women having
a chance to give their information. The sample size is small as compared to the entire population of fitness tracker user. It is bias and the result may not be accurate. 

### 3.3 Limitation of the Data
* A small sample size of 33 Fitbit user.
* Not enough information about the women


## 4. PROCESS
I chose R for data analysis because R is free, accessible and open source. R can perform data analysis, statistical analysis and create visualizations with 
lots of data. R can reproduce your analysis and create report in R Markdown to share with stakeholders. The daily_calories, daily_intensities and daily_steps data sets 
are included in the daily_activity. Also, we' ll exclude weight and heart rate data sets because the unique values for these data sets are small and not well represented.
We will use the following data set for analysis:
* dailyActivity_merged
* sleepDay_merged

```{r, include = FALSE}
#Install and load libraries
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
if(!require(gridExtra)) install.packages("gridExtra", repos = "http://cran.us.r-project.org")


library(tidyverse)
library(lubridate)
library(gridExtra)
```

```{r, include= FALSE}
#Load the daily_activity and daily_sleep data sets
daily_activity <- read_csv("./Bellabeat/dailyActivity_merged.csv")
daily_sleep <- read_csv("./Bellabeat/sleepDay_merged.csv")
```

### 4.1 Data Cleaning and Manipulation
We'll explore the data to understand the data structure. We'll clean and manipulate the data.

#### 4.1.1 Verify the number of FitBit Users
First, we will check for the number of users.
 
```{r}
#number of users for Id of daily_activity and daily_sleep
n_distinct(daily_activity$Id)
n_distinct(daily_sleep$Id)
```
There 33 Fitbit users who gave information about their daily activity and 24 Fitbit users gave information about their sleep. 

#### 4.1.2 Take a glimpse of the data sets
```{r}
#glimpse of daily_activity
glimpse(daily_activity)
```

We'll rename the ActivityDate to Date and convert the column from character strings to date values.

```{r}
#Rename ActivityDate and convert column from Character to date values
daily_activity <- daily_activity %>%
  rename(Date = ActivityDate) %>%
  mutate(Date = as_date(Date, format = "%m/%d/%Y"))
```


Let's get a glimpse of daily_sleep
```{r}
#Glimpse of sleep
glimpse(daily_sleep)
```
We'll rename the SleepDay to Date and convert the column from character strings to date values. Also we'll rename TotalTimeinBed to 
TotalMinutesInBed

```{r}
#Rename SleepDay, TotalTimeInBed and convert Character string to date values.
daily_sleep <- daily_sleep %>% 
  rename(Date = SleepDay) %>% 
  rename(TotalMinutesInBed = TotalTimeInBed) %>%
  mutate(Date = as.POSIXct(Date, format ="%m/%d/%Y %I:%M:%S %p" , tz=Sys.timezone()))
```

#### 4.1.3 Check for duplicates
```{r}
sum(duplicated(daily_activity))
sum(duplicated(daily_sleep))
```

There are 3 rows that are duplicated in the daily_sleep and none for daily_activity. We'll go ahead and see these rows and remove the duplicated rows.
```{r}
#Show entries with duplicates in daily_sleep
daily_sleep[duplicated(daily_sleep),]
```

```{r}
#Remove duplicated rows in daily sleep
daily_sleep <- daily_sleep[!duplicated(daily_sleep),]
```

```{r}
#Check again for duplicated rows 
sum(duplicated(daily_sleep))
```

#### 4.1.4 Check for missing values
We'll check for missing values in the data sets
```{r}
#Count the number of missing values in each column for daily_activity
colSums(is.na(daily_activity))
```
There are no missing values for daily_activity data set.

```{r}
#Count the number of missing values in each column for daily_sleep
colSums(is.na(daily_sleep))
```
There are no missing values for daily_sleep data set


### 5.ANALYZE
#### 5.1 Summary statistics of the data
Let's explore summary statistics of daily_activity.
```{r}
#summary statistics of daily_activity
daily_activity_summary <- daily_activity %>% select(TotalSteps, VeryActiveMinutes, FairlyActiveMinutes, 
                                     LightlyActiveMinutes, SedentaryMinutes, Calories) %>% summary()
daily_activity_summary
```

* The average steps in a day is 7638 which is less than 10,000 steps, the recommended steps most adult aimed to do.
* The average time for activity level varies across the levels.
* The average sedentary time is 991.2 which is approximately 16hrs in a day. Those who are sedentary spend more time being inactive.
* The average calories burned in a day is 2304.

We' ll explore the summary statistics of daily_sleep data. But first, we'll calculate the difference 
in time it takes  the fitbit user to fall asleep which is given by
$$
Difference\ in\ time = Total\ minutes \ in\ bed - Total\ minutes\
Asleep
$$


```{r}
#difference in time to fall asleep
daily_sleep <- daily_sleep %>% mutate(DifferenceInTime = TotalMinutesInBed - TotalMinutesAsleep)
```

```{r}
##summary statistics of daily_sleep_summary
daily_sleep_summary <- daily_sleep %>% select(TotalSleepRecords, TotalMinutesAsleep, TotalMinutesInBed, DifferenceInTime) %>% summary()
daily_sleep_summary
```
* The average number of sleep records a user get is approximately 1.This could mean that most user do not get enough sleep or a good quality sleep
* Also on average, the total minutes asleep is 419.5 which is approximately 7 hours of sleep.
* It also takes 39.31 minutes on average to fall asleep.

```{r}
#count Total Sleep Record
sleep_count <- daily_sleep %>% group_by(TotalSleepRecords) %>% summarize(sleep_count = n())
sleep_count
```
Users who had 1 record of sleep are 364, 2 records of sleep are 43 in number while 3 records of sleep are 3 in number. It is likely
that majority of the users don't get enough sleep that their bodies need.


### 6. SHARE
Let's make some visualizations to discover trends and relationships in the data set.

#### 6.1.  Daily calories burnt vs daily total steps
```{r}
#Correlation coefficient between daily total steps and daily Calories burnt
corr0 <- round(cor(daily_activity$TotalSteps, daily_activity$Calories, method = c("pearson")), 3)

p0 <- daily_activity %>% ggplot(aes(TotalSteps, Calories)) + geom_point(color = "red") + geom_smooth() +
                         geom_text(x = 2500, y = 4600, label = paste0('R = ', corr0), color = 'blue') +
                         labs(title = "Calories vs Total Steps")
p0
```
There is an upward trend between daily calories and daily steps. User tends to burn calories as steps increases. 
The relationship is positive and moderate with correlation coefficient of 0.592.

### 6.2 Calories burnt  vs  activity Level
```{r}
#Correlation coefficient between Very Active and Calories burnt
corr1 <- round(cor(daily_activity$VeryActiveMinutes, daily_activity$Calories, method = c("pearson")), 3)

#Scatter plot of Calories burnt vs Very Active
p1 <- daily_activity %>% ggplot(aes(VeryActiveMinutes, Calories)) + geom_point(color = "red") + geom_smooth()+
                        geom_text(x = 25, y = 4750,label = paste0('R = ', corr1),color = 'blue') +
                        labs(title = " Calories vs Very Active")


#Correlation coefficient between Fairly Active and Calories burnt
corr2 <- round(cor(daily_activity$FairlyActiveMinutes, daily_activity$Calories, method = c("pearson")), 3)

#Scatter plot of Calories burnt vs Fairly Active
p2 <- daily_activity %>% ggplot(aes(FairlyActiveMinutes, Calories)) + geom_point(color = "red") + geom_smooth() +
                         geom_text(x = 25, y = 4750,label = paste0('R = ', corr2),color = 'blue') +
                         labs(title = "Calories vs Fairly Active")
gridExtra::grid.arrange(p1, p2, nrow = 1 , ncol = 2)
```


```{r}
#Correlation coefficient between Lightly Active and Calories burnt
corr3 <- round(cor(daily_activity$LightlyActiveMinutes, daily_activity$Calories, method = c("pearson")), 3)

#Scatter plot of Calories burnt vs Lightly Active
p3 <- daily_activity %>% ggplot(aes(LightlyActiveMinutes, Calories)) + geom_point(color = "red") + geom_smooth() +
                         geom_text(x = 100, y = 4750,label = paste0('R = ', corr3), color ="blue") +
                         labs(title = "Calories vs Lightly Active ")

#Correlation coefficient between Sedentary and Calories burnt
corr4 <- round(cor(daily_activity$SedentaryMinutes, daily_activity$Calories, method = c("pearson")), 3)

#Scatter plot of Calories burnt vs Sedentary
p4 <- daily_activity %>% ggplot(aes(SedentaryMinutes, Calories)) + geom_point(color = "red") + geom_smooth() +
                         geom_text(x = 250, y = 4750,label = paste0('R = ', corr4), color = 'blue') +
                         labs(title = "Calories vs Sedentary")

gridExtra::grid.arrange(p3, p4, nrow = 1, ncol = 2)
```
From the graph, there is a sharp upward trend between daily calories burnt and very active users. A slight  upward trend for Lightly active user.
Also we can see a downward trend for users who are fairly active and sedentary . Users who are very active burn more calories than users who are 
lightly active fairly active and sedentary. The relationship between calories and activity level is much stronger for  
very active user with a correlation coefficient of 0.616 in comparison to the other users.

#### 6.3 Calories burnt vs sleep
Let's obtain the number of observations for daily_activity and daily_sleep
```{r}
#Number of observation for daily_activity
nrow(daily_activity)

#Number of observation for daily_sleep
nrow(daily_sleep)
```
So we have 940 records for daily_activity and 410 records for daily_sleep. We'll join daily_activity and daily_sleep by Id and Date 
using inner_join so we keep observations from daily_activity that have a matching key in daily_sleep
```{r}
#inner join for daily_activity and daily_sleep
daily_activity_sleep <- inner_join(daily_activity, daily_sleep, by=c ("Id", "Date"))
```

```{r}
corr5 <- round(cor(daily_activity_sleep$TotalMinutesAsleep, daily_activity_sleep$Calories, method = c("pearson")), 3)

#Scatter plot of Calories burnt vs Very Active
p5 <- daily_activity_sleep %>% ggplot(aes(TotalMinutesAsleep, Calories)) +
                               geom_point(color = "red") +
                               geom_smooth() +
                               geom_text(x = 125, y = 4750,label = paste0('R = ', corr5), color = 'blue') +
                               labs(title = "Calories vs Sleep")
p5
```

There is a slightly downward trend between daily calories burnt and number of minutes asleep. And majority of data points is concentrated around 400
which is approximately about 7hrs of sleep. At that 400 point, there are users whose calories burnt is low and some whose calories burnt is high.
In other words, though sleep plays a role in calories burnt, there are other factors that can also contribute to calories burnt.


#### 6.4 Sleep vs Activity level
```{r}
#Correlation coefficient between sleep and Very Active
corr6 <- round(cor(daily_activity_sleep$VeryActiveMinutes, daily_activity_sleep$TotalMinutesAsleep, method = c("pearson")), 3)

#Scatter plot of Sleep  vs Very Active
p6 <- daily_activity_sleep %>% ggplot(aes(VeryActiveMinutes, TotalMinutesAsleep)) + geom_point(color = "red") + geom_smooth()+
                               geom_text(x = 30, y = 750,label = paste0('R = ', corr1),color = 'blue') +
                               labs(title = "Sleep vs Very Active")


#Correlation coefficient between sleep and Fairly Active
corr7 <- round(cor(daily_activity_sleep$FairlyActiveMinutes, daily_activity_sleep$TotalMinutesAsleep, method = c("pearson")), 3)

#Scatter plot of Sleep vs Fairly Active
p7 <- daily_activity_sleep %>% ggplot(aes(FairlyActiveMinutes, TotalMinutesAsleep)) + geom_point(color = "red") + geom_smooth() +
                         geom_text(x = 25, y = 750,label = paste0('R = ', corr2),color = 'blue') +
                         labs(title = "Sleep vs Fairly Active")
gridExtra::grid.arrange(p6, p7, nrow = 1 , ncol = 2)
```


```{r}
#Correlation coefficient between Sleep and Lightly Active
corr8 <- round(cor(daily_activity_sleep$LightlyActiveMinutes, daily_activity_sleep$TotalMinutesAsleep, method = c("pearson")), 3)

#Scatter plot of Sleep  vs Lightly Active
p8 <- daily_activity_sleep %>% ggplot(aes(LightlyActiveMinutes, TotalMinutesAsleep)) + geom_point(color = "red") + geom_smooth()+
                               geom_text(x = 50, y = 750,label = paste0('R = ', corr8),color = 'blue') +
                               labs(title = "Sleep vs Lightly Active")


#Correlation coefficient between Sleep and Sedentary
corr9 <- round(cor(daily_activity_sleep$SedentaryMinutes, daily_activity_sleep$TotalMinutesAsleep, method = c("pearson")), 3)

#Scatter plot of Sleep vs Sedentary
p9 <- daily_activity_sleep %>% ggplot(aes(SedentaryMinutes, TotalMinutesAsleep)) + geom_point(color = "red") + geom_smooth() +
                         geom_text(x = 150, y = 750,label = paste0('R = ', corr9),color = 'blue') +
                         labs(title = "Sleep vs Sedentary")
gridExtra::grid.arrange(p8, p9, nrow = 1 , ncol = 2)
```
There is a slight upward trend for very Active users and a downward trend for fairly and sedentary. Users who are very active tend to sleep more 
than users who are Sedentary.


### 7. ACT
Based on the analysis, I recommend the following:

* Bellabeat should focus creating a reminder features in the app for sleep routines for users. Also the company should design and build routines
  practices like yoga, meditation and workouts that will enable users to fall asleep quickly since the average time users fall asleep is 39 minutes
  and improve their sleep quality.

* Create app in such a way it monitors users sleep activities when they sleep so that users can have a data of their sleep quality.

* Create notification to encourage sedentary and less active users to take more steps and be very active throughout the day so they can
  get good sleep and burn calories for those who want to lose weight.
  
* Create incentive contents for very active users to encourage them to keep a very active lifestyle.
